Vertical shear instability in astrophysical disks: vertically global linear calculations with thermal relaxation and applications to the Solar Nebula

### 2015-02-26 paper.tex
minor changes, new title
minor changes

\title{Linear analysis of the vertical shear instability in astrophysical disks}

\author{Min-Kai Lin, Andrew Youdin}
\affil{ $^1$Department of Astronomy and Steward Observatory, University of
  Arizona, 933 North Cherry Avenue, Tucson, AZ 85721, USA}
\email{minkailin@email.arizona.edu, youdin@email.arizona.edu}

\begin{abstract}
  Baroclinic astrophysical disks with equilibrium rotation profiles
  $\Omega$ that depends on the distance away from the disk midplane are
  potentially unstable to the Vertical Shear Instability (VSI). It is
  is important to understand the linear instability before
  considering the VSI as a hydrodynamic route to turbulent angular
  momentum transport in accretion disks. We study the linear,
  axisymmetric VSI in radially-local but vertically-global models
  of accretion disks. We include an energy equation with thermal
  relaxation on a characteristic timescale $t_c=\beta\Omega_k^{-1}$,
  where $\beta$ is a dimensionless parameter and $\Omega_k$ is the
  Keplerian rotation frequency, and explore how the VSI depends on the
  thermodynamic response  of the disk by varying $\beta$. We find
  increasing $\beta$ away from zero dramatically reduces VSI growth
  rates because this generally introduces a strongly stabilizing
  vertical entropy gradient. Our linear calculations are consistent
  with previous hydrodynamic simulations, which indicates the VSI 
  requires a short thermal relaxation timescale. For vertically  
  isothermal disks, we quantify `fast'
  thermal relaxation as $\beta < \epsilon|q|/(\gamma-1)$, where
  $\epsilon$ is the disk aspect-ratio, $q$ is the radial power-law
  index for the equilibrium temperature profile (which gives rise to vertical 
  shear) and $\gamma$ is the adiabatic index. We apply our linear
  results to a disk model of the Minimum Mass Solar Nebulae (MMSN), and
  find the MMSN is unstable to the VSI at tens of AU with a
  characteristic growth times of tens of local orbital periods. We 
  suggest the VSI is a viable mechanism for angular momentum
  transport in the outer parts of protoplanetary disks. 
\end{abstract}

@@

An important finding from these works is that the VSI requires a short
thermal timescale: perturbations need to be close to
isothermal, unless the disk is neutrally stratified in the vertical
direction. This is because vertical motions are stabilized by
buoyancy if the disk is sub-adiabatically stratified, as required for
convective stability. When the thermal timescale $t_c$ is parametrized
in units of the orbital period $P$, \cite{nelson13} find that
$t_c\lesssim 0.01P$ was needed to observe instability. That is, the
temperature profile of the disk essentially remains unchanged during
the disk evolution. This has been confirmed by 
\cite{stoll14}, who find the 
VSI is only sustained in radiation hydrodynamic simulations 
with the inclusion of heating from stellar irradiation.

>>>>

%\title{Linear analysis of the vertical shear instability in astrophysical disks}
\title{Vertical shear instability in astrophysical disks: vertically global linear calculations with thermal relaxation and applications to the Solar Nebula} 
\author{Min-Kai Lin, Andrew Youdin}
\affil{ $^1$Department of Astronomy and Steward Observatory, University of
  Arizona, 933 North Cherry Avenue, Tucson, AZ 85721, USA}
\email{minkailin@email.arizona.edu, youdin@email.arizona.edu}

\begin{abstract}
  Baroclinic astrophysical disks with equilibrium rotation profiles
  $\Omega$ that depends on the distance away from the disk midplane are
  potentially unstable to the Vertical Shear Instability (VSI). It is
  is important to understand the linear instability before
  considering the VSI as a hydrodynamic route to turbulent angular
  momentum transport in accretion disks. We study the linear,
  axisymmetric VSI in radially-local but vertically-global models
  of accretion disks. We include an energy equation with thermal
  relaxation on a characteristic timescale $t_c=\beta\Omega_k^{-1}$,
  where $\beta$ is a dimensionless parameter and $\Omega_k$ is the
  Keplerian rotation frequency, and explore how the VSI depends on the
  thermodynamic response  of the disk by varying $\beta$. We find
  increasing $\beta$ away from zero dramatically reduces VSI growth
  rates because this generally introduces a strongly stabilizing
  vertical entropy gradient. Our linear calculations are consistent
  with previous hydrodynamic simulations, which indicates the VSI 
  requires a short thermal relaxation timescale. For vertically  
  isothermal disks, we quantify `fast'
  thermal relaxation as $\beta < \epsilon|q|/(\gamma-1)$, where
  $\epsilon$ is the disk aspect-ratio, $q$ is the radial power-law
  index for the equilibrium temperature profile (which gives rise to vertical 
  shear) and $\gamma$ is the adiabatic index. We apply our linear
  results to a disk model of the Minimum Mass Solar Nebulae (MMSN), and
  find the MMSN is unstable to the VSI at tens of AU with a
  characteristic growth times of tens of local orbital periods. This 
  suggests the VSI is  dynamically important in the outer parts of protoplanetary disks. 
\end{abstract}

@@

An important finding from these works is that the VSI requires a short
thermal timescale: perturbations need to be close to
isothermal, unless the disk is neutrally stratified in the vertical
direction. This is because vertical motions are stabilized by
buoyancy if the disk is sub-adiabatically stratified, as required for
convective stability. When the thermal timescale $t_c$ is parametrized
in units of the orbital period $P_\mathrm{orb}$, \cite{nelson13} find that
$t_c\lesssim 0.01P_\mathrm{orb}$ was needed to observe instability. That is, the
temperature profile of the disk essentially remains unchanged during
the disk evolution. This has been confirmed by 
\cite{stoll14}, who find the 
VSI is only sustained in radiation hydrodynamic simulations 
with the inclusion of heating from stellar irradiation.

### 2015-02-26 setup.tex
minor changes

\subsection{Stability condition against axisymmetric adiabatic
  perturbations}\label{solberg}
A rotating flow is stable against axisymmetric adiabatic perturbations
if both Solberg-Hoiland criteria  are satisfied:
\begin{align}
  \kappa^2 - \frac{1}{\rho c_p}\nabla P \cdot \nabla S &> 0,\\
  -\frac{\p P}{\p z}\left(\frac{\p j^2}{\p r}\frac{\p S}{ \p z} -
    \frac{\p j^2}{\p z}\frac{\p S}{\p r} \right)&>0. \label{second_SH} 
\end{align}
\citep{tassoul78}, where $S = c_p\ln{P^{1/\gamma}/\rho}$ is the
specific entropy, $c_p$ is the heat capacity at constant
pressure (assumed to be constant), and $j\equiv r^2\Omega$ is the
specific angular momentum.

>>>>

\subsection{Stability condition against axisymmetric adiabatic
  perturbations}\label{solberg}
A rotating flow is stable against axisymmetric adiabatic perturbations
if both Solberg-Hoiland criteria  are satisfied:
\begin{align}
  \kappa^2 - \frac{1}{\rho c_p}\nabla P \cdot \nabla S &> 0,\\
  -\frac{\p P}{\p z}\left(\frac{\p j^2}{\p r}\frac{\p S}{ \p z} -
    \frac{\p j^2}{\p z}\frac{\p S}{\p r} \right)&>0. \label{second_SH} 
\end{align}
\citep{tassoul78}, where $S = c_p\ln{P^{1/\gamma}/\rho}$ is the
specific entropy, $c_p$ is the heat capacity at constant
pressure, and $j\equiv r^2\Omega$ is the
specific angular momentum.
